A Process for Emulating Comparative Oncology Trials with Real-world Evidence Studies (ENCORE): Process Development and Methodological Considerations for Oncology Real-World Data

### 2025-05-27 manuscript/manuscript.qmd
Added paragraph emulating oncology RCT estimands as per Bob's comments and Shirley's excellent additions

[**Analytic code sharing statement:**]{.underline} Code to reproduce this manuscript, including all figures and tables, can be found at <https://github.com/janickweberpals/encore-process-manuscript>.

@@

While most trial eligibility criteria could be operationalized in an automated fashion, our final criterion was an assessment of emulation feasibility and clinical relevance. This criterion involved extensive human review. The fundamental points considered in this step include an initial feasibility assessment of the data fitness,[@rivera2022friends] including assessment of whether critical eligibility criteria (e.g., biomarker status) including prognostic factors (e.g., ECOG performance score) are measurable and whether preliminary study size is reasonable. Lastly, trial candidates were ranked and shortlisted into primary and runner-up candidates based on their clinical and regulatory relevance.

A list of tentative, shortlisted primary candidates is presented in @tbl-rcts and the corresponding selection process is illustrated in the CONSORT diagrams (Supplementary Figures 1-4). Naturally, the majority of trials cover advanced (locally advanced, inoperable, recurrent/progressive disease) or metastatic cancer populations because a large proportion of drug development efforts have recently focused on these settings. A key objective that we aim to explore with the selected trials is to achieve a better understanding how different disease settings (early, late), line settings (\[neo\]adjuvant, first line, advanced lines of therapy), therapy protocols (monotherapy, combination therapy) and clinical characteristics (all-comer versus complex genetic or immunological signatures) can be emulated using RWD. If more thorough feasibility assessments suggest that the threat of bias from mis-measurement of key study parameters or residual confounding remains high, or that the study size is not sufficient, then runner-up candidate trials will be considered instead.

@@

@@

One important aspect when emulating oncology trials is the choice and estimation of the appropriate estimand of interest.[@rufibach2018] Particularly when emulating pivotal trials of paradigm-changing treatments, multiple aspects need to be considered such as the contemporaneity of the control cohort, the adoption rate of the novel medical product in routine care, the magnitude of the clinical treatment benefit and the rate in which (particularly patients in the control arm) discontinue or cross-over to the interventional treatment, which could lastly bias emulated treatment effects towards the null. To that end, comprehensive data explorations will be performed as part of the protocol development to contextualize these parameters and (if reported) draw comparisons to the emulated trial. Examples for such standard diagnostics are visualized in @fig-initiators. All exploratory analyses will be conducted blinded towards the outcome to avoid influencing study design and analytic choices.

@@

Another common challenge in the emulation of oncology trials is the estimation of an "intention-to-treat" (ITT) analogous estimand which is usually the primary estimand reported in oncology RCTs. Due to intercurrent events, such as non-adherence, crossover of a high proportion of patients from the control to the intervention arm, or differences in subsequent therapy lines there may be estimand differences between trial and emulation.[@manitz2022estimands] Although this is also a common challenge in the analysis of RCTs,[@rufibach2018] treatment practice in routine clinical might be less stringent compared to trials.[@merola2024calibrating] In order to derive comparable estimands it is therefore crucial to understand and contextualize the proportion and timing of treatment switching and discontinuation in both the trial and its emulation.

>>>>

[**Analytic code sharing statement:**]{.underline} Simulated examples and code to implement analytic workflows described in this manuscript are illustrated at <https://janickweberpals.github.io/imputation-ps-workflows/> and can be reproduced via the `encore.analytics` R package (<https://github.com/janickweberpals/encore.analytics/>).

@@

While most trial eligibility criteria could be operationalized in an automated fashion, our final criterion was an assessment of emulation feasibility and clinical relevance. This criterion involved extensive human review. The fundamental points considered in this step include an initial feasibility assessment of the data fitness,[@rivera2022friends] including assessment of whether critical eligibility criteria (e.g., biomarker status) including prognostic factors (e.g., ECOG performance score) are measurable, and whether preliminary study size based on number of patients observed in the data with the combination of treatment regimen and line of treatment is reasonable. Lastly, trial candidates were ranked and shortlisted into primary and runner-up candidates based on their clinical and regulatory relevance.

A list of tentative, shortlisted primary candidates is presented in @tbl-rcts and the corresponding selection process is illustrated in the CONSORT diagrams (Supplementary Figures 1-4). Naturally, the majority of trials cover advanced (locally advanced, inoperable, recurrent/progressive disease) or metastatic cancer populations because a large proportion of drug development efforts have recently focused on these settings. A key objective that we aim to explore with the selected trials is to achieve a better understanding how different disease settings (early, late), line settings (\[neo\]adjuvant, first line, advanced lines of therapy), therapy protocols (monotherapy, combination therapy) and clinical characteristics (all-comer versus complex genetic or immunological signatures) can be emulated using RWD. If more thorough feasibility assessments after applying all relevant eligibility criteria suggest that the threat of bias from mis-measurement of key study parameters or residual confounding remains high, or that the study size is not sufficient, then runner-up candidate trials will be considered instead.

@@

**Emulating the RCT estimand**

One important aspect when emulating oncology trials is the choice and estimation of the appropriate estimand of interest.[@rufibach2018] In prior RCT-DUPLICATE efforts, a “while on treatment” strategy was chosen for database studies that were designed to emulate trials where the primary trial analysis was intention-to-treat (e.g., a “treatment policy” estimand). This estimand was chosen to mimic the high adherence typically observed in large cardiovascular clinical trials in the context of low adherence in clinical practice. However, a “while on treatment” estimand would be inappropriate when emulating oncology trials, not only because of highly informative censoring, but also because OS outcome analyses in oncology are typically “treatment policy” estimands that allow crossover.[@manitz2022estimands] We plan to focus on “treatment policy” estimands for the primary analyses in our trial emulations. However, allowing crossover can dilute treatment effects, moving point estimates toward the null. Therefore, differences in cross-over and discontinuation rates between the trial and emulation may still contribute to estimand differences. Recognizing these practical challenges, we will carefully characterize the patterns of discontinuation or cross-over in both the trials and the database studies designed to emulate them.

@@

Particularly when emulating pivotal trials of paradigm-changing treatments, multiple aspects need to be considered, such as the contemporaneity of the control cohort, the adoption rate of the novel medical product in routine care, the magnitude of the clinical treatment benefit and the rate in which patients discontinue or cross-over, which could dilute effect estimates.

To that end, comprehensive data explorations will be performed as part of the protocol development to contextualize these parameters and (if reported) draw comparisons to the emulated trial. Examples for such standard diagnostics are visualized in @fig-initiators. All exploratory analyses will be conducted blinded towards the outcome to avoid influencing study design and analytic choices.

@@

Another common challenge in the emulation of oncology trials is the estimation of an "treatment policy"/"intention-to-treat" analogous estimand which is usually the primary estimand reported in oncology RCTs. Due to intercurrent events, such as non-adherence, crossover of a high proportion of patients from the control to the intervention arm, or differences in subsequent therapy lines there may be estimand differences between trial and emulation.[@manitz2022estimands] Although this is also a common challenge in the analysis of RCTs,[@rufibach2018] treatment practice in routine clinical practice might be less stringent compared to trials.[@merola2024calibrating] In order to derive comparable estimands it is therefore crucial to understand and contextualize the proportion and timing of treatment switching and discontinuation in both the trial and its emulation.
